---
title: "Tutorial - Plotting data from cultures"
author: "Daniel Vaulot"
date: '`r format(Sys.time(), "%d %m %Y")`'
header-includes:
   - \usepackage{color, fancyvrb}
output:
  rmdformats::readthedown:
    highlight: kate
    number_sections : yes    
  pdf_document: 
    toc: yes
    toc_depth: 2
    number_sections : yes
---

```{r knitr_init, echo=FALSE, warning=FALSE, cache=TRUE}
library(knitr)
library(rmdformats)
library("kableExtra")

## Global options
# The following line is necessary to make sure that 
# everything is not reevaluated each time the file is knit
# Note : in the case of this report it is necessary to leave cache= FALSE

options(max.print="75")
knitr::opts_chunk$set(fig.width=8, 
                      fig.height=6, 
                      eval=TRUE, 
                      cache=TRUE,
                      echo=TRUE,
                      prompt=FALSE,
                      tidy=TRUE,
                      comment=NA,
                      message=FALSE,
                      warning=FALSE)
opts_knit$set(width=85)
```


# Aim

This tutorial explain how plot growth curves for phytoplankton cultures in different conditions.
* Antibiotics treatment : 4 RCC cultures, 8 days, 2 antibiotics, 5 concentrations, 


# Downloads

Install the following software :  


* R studio : https://www.rstudio.com/products/rstudio/download/#download

* Download and install the following libraries by running under R studio the following lines

```R
install.packages("dplyr")     # To manipulate dataframes
install.packages("tidyr")     # To manipulate dataframes

install.packages("stringr")   # To strings

install.packages("ggplot2")   # for high quality graphics
install.packages("gridExtra") # for grids

install.packages("plotrix" )  # needed for standard error


```

# Data used

* **cultures antibiotics.txt** contains data obtained by Priscilla Gourvil on antibiotics treatment of RCC strains
* **grazing experiment.xlsx** contains data obtained by Valeria Jimenez on grazing experiment on Micromonas


# Tutorial description

## Load the libraries

```{r, results='hide', message=FALSE, warning=FALSE}
library("ggplot2")
library("gridExtra")
library("plotrix" ) # needed for standard error
library("dplyr")
library("tidyr")
library("stringr")
library("readxl")
```

## Antibiotics treatments
### Read and reformat the data
Read the data
```{r}
cell<- read.table("cultures antibiotics.txt", header=TRUE, sep="\t", na.strings="NA", dec=".", strip.white=TRUE)
knitr::kable(head(cell))
```

Change from wide format to long format
```{r}
cell<- gather(cell, X1:X7, key = "day", value = "cell_number") 
knitr::kable(head(cell))
```

Reformat day as numeric
```{r}
cell$day<-as.numeric(str_replace(cell$day, "X", ""))
knitr::kable(head(cell))
```

Reformat concentration as character
```{r}
cell$Concentration<-as.character(cell$Concentration)
knitr::kable(head(cell))
```

Compute mean and SD for each RCC, Antibio, Concentration and day using dplyr
```{r}
cell_1<- cell %>% group_by(RCC,Antibio,Concentration, day)%>%
         summarise (cell_mean=mean(cell_number),cell_sd=sd(cell_number), cell_se=std.error(cell_number))
knitr::kable(head(cell_1))
```

### Define graphics options
Define the color, line type and symbol shape
```{r}
Concentration_color<-c("0.2"="white","0.5"="white", "0.8"="white", "1"="black", "1.5"="black", "2"="black")
Concentration_linetype<-c("0.2"=1,"0.5"=1, "0.8"=1, "1"=2, "1.5"=2, "2"=2)
Concentration_shape<-c("0.2"=21,"0.5"=22, "0.8"=21, "1"=22, "1.5"=21, "2"=22)
```

Define graphics options
```{r}
scaling_factor=15
cell_label <- expression (paste("cell.",mL^-1))
cell_breaks=c(100,1000,10000,100000,1000000)
x_max=8
x_breaks=c(0, 2,4,6,8)
x_labels=c("0", "2","4","6","8")
```

\pagebreak

### Plot the data
```{r, fig.height = 15, fig.width=10}
plot1<- ggplot(cell_1, aes(x=day, y=cell_mean, group = Concentration, xmin=0, xmax=x_max, 
               shape=Concentration, fill=Concentration, linetype=Concentration)) + 
              facet_wrap(~ RCC + Antibio, nrow=4, ncol=2, scales="free") +
	            geom_line (size=0.8, colour="black") + 
              geom_point(size = 4) + 
              geom_errorbar(aes(ymin=cell_mean-cell_se, ymax=cell_mean+cell_se),width=0.2, linetype=1) +
              theme_bw(scaling_factor) + 
              theme(panel.border = element_rect(colour = "black"), panel.grid.major = element_blank(), panel.grid.minor = element_blank(),	
              axis.line = element_line(colour = "black"), 
		          legend.title=element_text(size=scaling_factor), legend.key=element_blank(),
		          axis.title = element_text(size=scaling_factor), 
		          legend.text=element_text(size=scaling_factor), legend.key.height = unit(1, "cm"), 
		          axis.text = element_text(size=0.8*scaling_factor), panel.background = element_rect(fill="white")) + 
	            theme(legend.position = "top", legend.box = "horizontal")  +
	            labs(x = "Days", y = cell_label ) +
              scale_x_continuous(breaks=x_breaks, labels=x_labels) + 
	            scale_y_log10(breaks = cell_breaks ,labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
              annotation_logticks(sides = "lr") + 
	            scale_fill_manual(values=Concentration_color) + 
              scale_shape_manual(values=Concentration_shape) + 
              scale_linetype_manual(values=Concentration_linetype)

# Add next line to zoom		
# + coord_cartesian(ylim=c(100, 10000000))		

plot1

# Next can be used to save the plot as pdf		
# ggsave(file="Fig 1 version 2.0.pdf", plot=plot1, scale=5, width = 7, height = 10, units = "cm", useDingbats=FALSE)

```


## Grazing experiment

Micromonas are fed with fluorescent labelled beads and one looks at the % of cells that have beads
The idea is to do a plot with 2 different scales for the y axis.

#### Read the data

```{r}
 grazing <- read_xlsx("grazing experiment.xlsx", sheet="RCC2306")

# Compute a new variable with the same scale as the cell concentration to be able to plot on the same graph
 grazing$cell_beads_pct_scaled <- grazing$cell_beads_pct*200000

  knitr::kable(head(grazing))
```

#### Plot the data
Demonstrate the use of sec_axis.
```{r}
plot <- ggplot(data=grazing, aes(x=time, y=cell_ml, ymin=0, fill=treatment)) + 
              geom_point(size = 4, shape = 21) +
	            labs(x = "Time (hours)", y = "Cells per mL - circles" ) +
              geom_point(size = 4, shape = 22, aes(x=time, y=cell_beads_pct_scaled, fill=treatment)) +
              scale_y_continuous(sec.axis = sec_axis(~./200000, name="% of cells with beads - squares"))
print(plot)
```



