---
title: "Tutorial - Maps with R"
author: "Daniel Vaulot"
date: '`r format(Sys.time(), "%d %m %Y")`'
always_allow_html: yes
header-includes:
   - \usepackage{xcolor, fancyvrb}
   - \usepackage{hyperref} 
   - \hypersetup{colorlinks=true, citecolor=darkgray, linkcolor=darkgray} 
output:
  pdf_document: 
    toc: yes
    toc_depth: 2
    number_sections : yes
    fig_height : 6
    fig_width : 7

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim

This tutorial explain how to make maps with different packages under R


# Downloads

Install the following software :  


* R studio : https://www.rstudio.com/products/rstudio/download/#download


Download and install the following libraries by running under R studio the following lines

```R
install.packages("dplyr")     # To manipulate dataframes
install.packages("stringr")   # To manipulate strings
install.packages("readxl")    # To read directly the data from an Excel file

install.packages("ggplot2")   # for high quality graphics
install.packages("leaflet")

```

# References

* Leaflet for R : https://rstudio.github.io/leaflet/


# Tutorial description

## Load necessary libraries

```{r, results='hide', message=FALSE}
  library("readxl")
  library("leaflet")
```


## Use of leaflet for R

### Simple map
```{r}
m <- leaflet(options = leafletOptions()) %>%
  addTiles() %>%  # Add default OpenStreetMap map tiles
  addMarkers(lng=-3.9375, lat=48.77777778, popup="Roscoff Astan station") %>%
  setView (lng=-3.95, lat=48.75, zoom=12)
m  # Print the map

```
\includegraphics[width=300px]{img/leaflet_astan.png}


### Maps of number of cultures isolated from different stations of the Roscoff Culture Collection (RCC) 
Circle radius is proportionnal to number of cultures
* addCircleMarkers: size is in pixel and therefore circle size does not change when zooming
* addCircles: size is in meters and change when zooming


```{r}
df <- read_excel("RCC_cultures_per_station.xlsx")

m <-  leaflet(options = leafletOptions()) %>%
      addTiles() %>%  # Add default OpenStreetMap map tiles 
      addCircleMarkers(data = df, lat = ~ Latitude, lng = ~ Longitude, radius = ~n_cult/10, color = "red", fill = TRUE)

m
```
\includegraphics[width=300px]{img/leaflet_circle_markers.png}

```{r}
m <-  leaflet(options = leafletOptions()) %>%
      addTiles() %>%  # Add default OpenStreetMap map tiles 
      addCircles(data = df, lat = ~ Latitude, lng = ~ Longitude, radius = ~n_cult*1000, color = "red", fill = TRUE)

m
```
\includegraphics[width=300px]{img/leaflet_circles.png}


### Change the background of the maps

See http://leaflet-extras.github.io/leaflet-providers/preview/index.html

* Ocean Base map

```{r}
m <- m %>% addProviderTiles(providers$Esri.OceanBasemap)
m
```
\includegraphics[width=300px]{img/leaflet_ocean_basemap.png}


* Esri World Imagery

```{r}
m <- m %>% addProviderTiles(providers$Esri.WorldImagery)
m
```

\includegraphics[width=300px]{img/leaflet_esri.png}

